---
title: "penguin_assessment.r"
output: html_document
---

# Question 4

### *Loading the data and relevant packages*
```{r}
#First make sure to set your working directory using setwd()
library(palmerpenguins)
library(ggplot2)
library(dplyr)
library(janitor)
library(tidyverse)
library(ragg)
```


### *Cleaning the data*
```{r}
#Reading in the raw data
penguins_raw <- read.csv("data_raw/penguins_raw.csv")
# Defining the cleaning function
cleaning <- function(data_raw){
  data_raw %>%
    clean_names() %>%
    remove_empty(c("rows", "cols")) %>%
    select(-starts_with("delta")) %>%
    select(-starts_with("x")) %>%
    select(-comments)
}
#Making another function to subset the data and remove empty rows in body mass
remove_empty_body_mass <- function(data_clean){
  data_clean %>%
    filter(!is.na(body_mass_g)) %>%
    select(species, body_mass_g, culmen_length_mm)
}
#Making another function to subset the data and remove empty rows in culmen length
remove_empty_culmen_length <- function(data_clean){
  data_clean %>%
    filter(!is.na(culmen_length_mm)) %>%
    select(species, body_mass_g, culmen_length_mm)
}
# Creating a new function and saving it to the cleaning file
cleaning_new <-function(data_raw){
  data_raw %>%
    cleaning() %>%
    remove_empty_body_mass() %>%
    remove_empty_culmen_length()
}
# Calling on the function
penguins_clean <- cleaning_new(penguins_raw)
# Checking the results
names(penguins_clean)
# Saving the clean data as penguins_clean.csv in a new folder called data_clean 
write.csv(penguins_clean, "data_clean/penguins_clean.csv")
```

### *Running a statistical test and plotting the data*

Linear regression model between culmen length and body mass


##### Plotting a scatterplot
```{r}
ggplot(penguins_clean, 
       aes(x=body_mass_g, 
           y=culmen_length_mm)) +
  geom_point(size=3, alpha=0.8) +
  theme_bw() +
  labs(title="Body Mass vs Culmen Length", 
       subtitle = "Culmen length and body mass of penguins",
       x="Body mass (g)",
       y = "Culmen length (mm)")
```

##### Fitting a regression line onto the plot

```{r}
ggplot(penguins_clean, 
       aes(x=body_mass_g, 
           y=culmen_length_mm)) +
  geom_point(size=3, alpha=0.8) +
  theme_bw() +
  labs(title="Body Mass vs Culmen Length", 
       subtitle = "Culmen length and body mass of penguins",
       x="Body mass (g)",
       y = "Culmen length (mm)")+
  geom_smooth(method="lm")
```


##### Regression model analysis
```{r}
#creating a regression model
penguins_mod <- lm(culmen_length_mm~body_mass_g, penguins_clean)
#penguins_mod_quad <-lm(culmen_length_mm~body_mass_g+ I(body_mass_g^2), penguins_clean)
summary(penguins_mod)
anova(penguins_mod)
```
From the summary output, we can work out that the equation of the regression line is y = (4.05 x 10^-3)x + (2.69 x 10^1). Because our p-value is less than 0.5, it suggests that we can use body mass to predict culmen length. But from the multiple R-squared value, we can see that the linear model only explains 35% of the variation in the data. To investigate this further, I will check if the data violates any assumptions of linear regression.  

##### Checking assumptions

###### ... for normal distribution
```{r}
#using QQ plot
plot(penguins_mod, which=2)
#using a histogram
ggplot(penguins_clean, aes(x=log(body_mass_g))) + geom_histogram()
ggplot(penguins_clean, aes(x=culmen_length_mm)) + geom_histogram()
```
From the plot you can see that the data is roughly normally distributed.

###### ... for equal variance
```{r}
plot(penguins_mod, which=1)
```

We can see that the data is not equally variable because there seems to be a decrease in vairance of the residuals from left to right. 

Even though the data doesn't match assumptions of linear regression perfectly, it is still better than the quadratic nad other polynomial models, so we can still map a linear model onto it to test for causality.

##### Simpson's paradox - looking at individual species
An alternative reason for the low R-squared value is Simpson's paradox. This is a phenomenon in which a trend a trend appearing in several groups of data disappears when the groups are pooled. To test this, i will split up the data according to species and individually test for linear regression. 


###### Plotting a scatterplot highlighting species
```{r}
ggplot(penguins_clean, 
       aes(x=body_mass_g, 
           y=culmen_length_mm, 
           colour = species, 
           shape = species)) +
  geom_point(size=3, alpha=0.8) +
  theme_bw() + 
  scale_colour_manual(values=c("darkorange", "purple", "cyan4")) +
  labs(title="Body Mass vs Culmen Length", 
       subtitle = "Culmen length and body mass for Adelie, Chinstrap and Gentoo penguins",
       x="Body mass (g)",
       y = "Culmen length (mm)",
       colour="Penguin species",
       shape = "Penguin species") + 
  geom_smooth(method="lm", se=FALSE)
```

From this we can see that Chinstrap penhuins don't fit the general linear trend of the pooled dataset. 

###### Linear regression model analysis on Chinstrap penguins

```{r}
chinstrap <- filter(penguins_clean, species=="Chinstrap penguin (Pygoscelis antarctica)")
chinstrap
chinstrap_mod <- lm(culmen_length_mm ~ body_mass_g, chinstrap)
summary(chinstrap_mod)
anova(chinstrap_mod)
```
R-squared for this regression line is 0.26, and the p-value is less than 0.05. So, for chinstraps we can use body mass to predict culmen length. 

###### Linear regression model analysis on Gentoo penguins
```{r}
gentoo <- filter(penguins_clean, species=="Gentoo penguin (Pygoscelis papua)")
gentoo
gentoo_mod <- lm(culmen_length_mm ~ body_mass_g, gentoo)
summary(gentoo_mod)
anova(gentoo_mod)
```
R-squared for this regression line is 0.45, and the p-value is less than 0.05. So, for gentoos we can use body mass to predict culmen length. 

###### Linear regression model analysis on Adelie penguins 
```{r}
adelie <- filter(penguins_clean, species=="Adelie Penguin (Pygoscelis adeliae)")
adelie
adelie_mod <- lm(culmen_length_mm ~ body_mass_g, adelie)
summary(adelie_mod)
anova(adelie_mod)
```
R-squared for this regression line is 0.3, and the p-value is less than 0.05. So, for adelies we can use body mass to predict culmen length. 

###### Putting it altogether without Chinstraps
```{r}
#binding adelie and gentoo dataframes
penguins_AG <- rbind(adelie, gentoo)
#plotting a regression line through the penguin_AG data
penguins_AG_plot <- ggplot(penguins_AG, 
    aes(x=body_mass_g, 
           y=culmen_length_mm)) +
  geom_point(size=3, alpha=0.8) +
  theme_bw() +
  labs(title="Body Mass vs Culmen Length", 
       subtitle = "Culmen length and body mass for Adelie and Gentoo penguins",
       x="Body mass (g)",
       y = "Culmen length (mm)")+
  geom_smooth(method="lm")
penguins_AG_plot
#fitting a linear regression model to this
penguins_AG_mod <- lm(culmen_length_mm ~ body_mass_g, penguins_AG)
summary(penguins_AG_mod)
anova(penguins_AG_mod)
```
Now, without the Chinstrap data, we can see that the linear model accounts for 77% of the variation in the data (instead of 35% with Chinstraps), and the p-value is also below 0.05. This indicates that we can use body mass to predict culmen length. 

### *Saving the figure*
Firstly, I made functions to plot the figure and then to save the plot as a .png

```{r}
## Function to plot the body mass and culmen length data
plot_M_vs_CL_figure <- function(penguins_AG){
  penguins_AG %>%
    ggplot(aes(x=body_mass_g, 
           y=culmen_length_mm)) +
  geom_point(size=3, alpha=0.8) +
  theme_bw() +
  labs(title="Body Mass vs Culmen Length", 
       subtitle = "Culmen length and body mass for Adelie and Gentoo penguins",
       x="Body mass (g)",
       y = "Culmen length (mm)")+
  geom_smooth(method="lm")
}
penguins_AG_plot<- plot_M_vs_CL_figure(penguins_AG)
penguins_AG_plot
## Function to save the plot as .png
save_LM_plot_png <- function(penguins_AG, filename, size, res, scaling){
  agg_png(filename, width = size, 
                    height = size, 
                    units = "cm", 
                    res = res, 
                    scaling = scaling)
  penguins_AG_plot <- plot_M_vs_CL_figure(penguins_AG)
  print(penguins_AG_plot)
  dev.off()
}
# Call the function that will save it as a .png
save_LM_plot_png(penguins_AG, "figures/penguins_AG_plot.png", 
                      size = 15, res = 600, scaling = 1)
```
